Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-19 paper.tex
...

Pingus%
\footnote{Pingus: \url{http://pingus.seul.org/}}
is an open-source clone of Lemmings%
\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
a puzzle-platformer video game.      
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit%
\footnote{Pingus gameplay: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus}
\caption{Pingus gameplay.
\label{fig.sweeney}
}
\end{figure}

Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase is about 40.000 lines of code (LoCs)%
\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}, divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

According to Tim Sweeney (of Unreal Engine fame), about half the complexity in
game development resides in \emph{simulation} (aka \emph{game logic}), but
which accounts for only 10\% of the CPU budget~\cite{games.sweeney}.
The game logic ``models the state of the game world as interacting objects
evolve over time''.
The high development costs contrasting with the low impact on performance
appeals for alternatives with productivity in mind, especially considering that
it is the game logic that varies the most between projects.
Sweeney states that ``will gladly sacrifice 10\% of our performance for 10\%
higher productivity''.

@@

This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
<!--, helping on encapsulation and cohesion.-->
Nonetheless, the lowest ratio numbers above correlate to the parts of the game
logic that we consider more susceptible to structured reactive programming.
For instance, the *Pingu* behavior (*ratio 0.80*) contains complex animations
that are affected by timers, game rules, and user interaction.
In contrast, the *Option* screen (*ratio 0.97*) is a simple UI grid with
trivial mouse interactions.

%[cpp_uncompressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP
%[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
%[ceu_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL

@@

The rewriting process consisted of identifying sets of callbacks implementing
*control-flow behaviors* in the game and translating them to \CEU using
appropriate structured constructs.
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{double-click-opt}
\caption{Double click detection.
\label{fig.armageddon}
}
\end{figure}

In Pingus, a double click in the \emph{Armageddon button} at the bottom right
of the screen literally explodes all pingus (Figure~\ref{fig.armageddon}).
Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-anim}
\caption{ The \emph{Bomber action}.
\label{fig.bomber.anim}
}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-state}
\caption{ State machine for the *Bomber* animation.
\label{fig.bomber.state}
}
\end{figure}

The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius (Figure~\ref{fig.bomber.anim}).
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber.state}) with actions associated to specific frames as
follows:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

TODO
%*([This video][youtube_bomber] presents the sound effects.)*
%[youtube_bomber]: https://youtu.be/QLXIT59il6o?t=306

\begin{figure*}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Bomber::Bomber (Pingu* p) :
    <...>
    sprite(<...>),              // bomber sprite
    sound_played(false),        // tracks state 2               @def_1
    particle_thrown(false),     // tracks state 3
    colmap_exploded(false),     // tracks state 3
    gfx_exploded(false)         // tracks state 4               @def_2
{
    <...>
    // 1. 0th frame: plays a "Oh no!" sound.
    get_world()->play_sound("ohno", pingu->get_pos());          @sound_ohno
}

void Bomber::update () {                                        @update_ini
    sprite.update();                                            @update_1
    <...>   // pingu movement                                   @update_2

    // 2. 10th frame: plays a "Bomb!" sound.                    @sound_bomb_1
    if (sprite.get_current_frame()==10 && !sound_played) {
        sound_played = true;
        get_world()->play_sound("plop", pingu->get_pos());
    }                                                           @sound_bomb_2

    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
    if (sprite.get_current_frame()==13 && !particle_thrown) {   @state_3_1
        particle_thrown = true;
        get_world()->get_pingu_particle_holder()->add_particle(...);
    }
    if (sprite.get_current_frame()==13 && !colmap_exploded) {
        colmap_exploded = true;
        get_world()->remove(bomber_radius, <...>);
    }                                                           @state_3_2

    // 5. Last frame: kills the Pingu
    if (sprite.is_finished ()) {                                @die_1
        pingu->set_status(Pingu::PS_DEAD);
    }                                                           @die_2
}                                                               @update_end

void Bomber::draw (SceneContext& gc) {
    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
    // 4. Game tick: hides the explosion sprite
    if (sprite.get_current_frame()==13 && !gfx_exploded) {      @state_3_3
        gfx_exploded = true;                                    @state_3_4
        gc.color().draw(explo_surf, <...>);                     @state_4
    }
    gc.color().draw(sprite, pingu->get_pos());
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> _ActionName__Enum
do
    <...>
    spawn Mover();                          // spawn the pingu movement to execute in the "background"      @move

    var&? Sprite s = spawn Sprite(<...>);   // spawn the bomber animation to execute in the "background"    @sprite
    watching s do                                                   @watch-1
        // 1. 0th frame: plays a "Oh no!" sound.                    @anim_1
        {Sound::PingusSound::play_sound("ohno", 0.5, 0.0)};

        // 2. 10th frame: plays a "Bomb!" sound.
        await outer.game.dt until s!.sprite.frame == 10;            @frame_1
        {Sound::PingusSound::play_sound("plop", 0.5, 0.0)};

        // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
        await outer.game.dt until s!.sprite.frame == 13;            @frame_2
        spawn PinguParticles(<...>) in outer.pingu_particles;       @particles
        call Game_Remove({&bomber_radius}, <...>);
        do                                                          @do
            <...>
            spawn Sprite(<...>);            // explosion            @explo

            // 4. Game tick: hides the explosion sprite
            await outer.game.dt;                                    @frame_3
        end                                                         @end
        await FOREVER;
    end                                                             @watch-2

    // 5. Last frame: kills the pingu
    escape {ActionName::DEAD};                                      @anim_2
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Advancing pages in the \emph{Story Screen}.
\label{lst.story}
}
\end{figure*}

In C++, the class \code{Bomber} defines the callbacks \code{draw} and
\code{update} to manage the state machine described above:
%
The class first defines one state variable for each action to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) updates the pingu animation and
movement on every frame regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
switches to \emph{state-2} (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 25--32 and
43--44).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 35--37).
%
Note that a single numeric state variable suffices to track the states, but the
original authors probably chose to encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., *methods + mutable state*).
%
Like the C++ implementation for detecting double clicks, note also that the
state machine is encoded across 3 different methods, each intermixing code with
unrelated functionality.

The equivalent code for the \emph{Bomber action} in \CEU doesn't require state
variables and reflects the sequential state machine implicitly, using
\code{await} statements in direct style to separate the actions.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine~\cite{sync_async.cooperative}: a subroutine that retains runtime
state, such as local variables and the program counter, across reactions to
events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other abstractions
and execute in separate through the \code{spawn} primitive (ln. 4,6).
The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every frame.
%
The code tracks the animation instance (ln. 7--27), performing the last bomber
action on termination (ln. 30).
As soon as the animation starts, the code performs the first action (ln. 8).
The intermediate actions are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all actions happen in a contiguous
chunk of code (ln. 6--30) which handles no extra functionality.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{story-anim}
\caption{The \emph{Story screen}.
\label{fig.story}
}
\end{figure}

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens (Figure~\ref{fig.story}).
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.

@@

@@

\textbf{Case Study: Bomber `draw` and `update` callbacks}

@@

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
animation frames (Figure~\ref{fig.TODO}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

@@

\textbf{Case Study: The Pingus Container}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus_create_die-anim}
\caption{Creation and death of pingus.
\label{fig.pingus_create_die}
}
\end{figure}

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude
(Figure~\ref{pingus_create_die}).

@@

code/await Pingu (<...>) do
    <...>
    loop do
        await game.dt;
        if Pingu_Is_Out_Of_Screen() then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen spawns a new \code{Pingu} on every invocation of
\code{Pingu\_Spawn}.
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its *natural termination*, the instance is automatically removed
from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{options-anim-opt}
\caption{The \emph{Mouse Grab} configuration option.
\label{fig.options}
}
\end{figure}

The *Mouse Grab* option restricts the mouse movement to the game window
boundaries (Figure~\ref{fig.options}).
The option can be set anywhere in the game by pressing *Ctrl-G*.
In addition, the *Options* menu has a check box to toggle the *Mouse Grab*
option with mouse clicks while still responding to *Ctrl-G* presses.

@@

\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus}
\caption{Pingus gameplay.
\label{fig.pingus}
}
\end{figure}

Pingus is an open-source clone of Lemmings%
\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
a puzzle-platformer video game (Figure~\ref{fig.pingus}).      
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit%
\footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
%
Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase%
\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}
is about 40.000 lines of code (LoCs), divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

According to Tim Sweeney (of Unreal Engine fame), about half the complexity in
game development resides in \emph{simulation} (aka \emph{game logic}), but
which accounts for only 10\% of the CPU budget~\cite{games.sweeney}
(Figure~\ref{fig.sweeney}).
The game logic ``models the state of the game world as interacting objects
evolve over time''.
The high development costs contrasting with the low impact on performance
appeals for alternatives with productivity in mind, especially considering that
it is the game logic that varies the most between projects.
Sweeney states that ``will gladly sacrifice 10\% of our performance for 10\%
higher productivity''.

@@

\footnote{ Game logic code: \url{https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP} }.
%[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
%[ceu_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL

This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
<!--, helping on encapsulation and cohesion.-->
Nonetheless, the lowest ratio numbers above correlate to the parts of the game
logic that we consider more susceptible to structured reactive programming.
For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
animations that are affected by timers, game rules, and user interaction.
In contrast, the \emph{Option screen} (\emph{ratio 0.97}) is a simple UI grid
with trivial mouse interactions.

@@

The rewriting process consisted of identifying sets of callbacks implementing
\emph{control-flow behaviors} in the game and translating them to \CEU using
appropriate structured constructs.
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.

@@

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{double-click-opt}
%\caption{Double click detection.
%\label{fig.armageddon}
%}
%\end{figure}
%(Figure~\ref{fig.armageddon}).

In Pingus, a double click in the \emph{Armageddon button} at the bottom right
of the screen literally explodes all pingus.%
\footnote{Double click animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#1} }
Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.

@@

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{bomber-anim}
%\caption{ The \emph{Bomber action}.
%\label{fig.bomber.anim}
%}
%\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-state}
\caption{ State machine for the \emph{Bomber animation}.
\label{fig.bomber}
}
\end{figure}

\begin{figure*}[th!]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Bomber::Bomber (Pingu* p) :
  <...>
  sprite(<...>),          // bomber sprite
  sound_played(false),    // tracks state 2
  particle_thrown(false), // tracks state 3
  colmap_exploded(false), // tracks state 3
  gfx_exploded(false)     // tracks state 4
{
  <...>
  // 1. plays a "Oh no!" sound.
  get_world()->play_sound("ohno");
}

void Bomber::update () {
  sprite.update();
  <...>   // pingu movement

  // 2. plays a "Bomb!" sound.
  if (sprite.frame()==10 && !sound_played) {
    sound_played = true;
    get_world()->play_sound("plop");
  }

  // 3. particles, terrain, explosion sprite
  if (sprite.frame()==13 && !particle_thrown) {
    particle_thrown = true;
    get_world()->get_particles()->add(...);
  }
  if (sprite.frame()==13 && !colmap_exploded) {
    colmap_exploded = true;
    get_world()->remove(bomber_radius, <...>);
  }

  // 5. kills the Pingu
  if (sprite.is_finished ()) {
    pingu->set_status(PS_DEAD);
  }
}

void Bomber::draw (SceneContext& gc) {
  // 3. particles, terrain, explosion sprite
  // 4. tick: hides the explosion sprite
  if (sprite.frame()==13 && !gfx_exploded) {
    gfx_exploded = true;
    gc.color().draw(explo_surf, <...>);
  }
  gc.color().draw(sprite, pingu->get_pos());
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName
do
  <...>
  spawn Mover();  // movement in the background
  var&? Sprite s = spawn Sprite(<...>);
                  // animation in the background
  watching s do
    // 1. plays a "Oh no!" sound.
    {Sound::PingusSound::play_sound("ohno")};

    // 2. plays a "Bomb!" sound.
    await game.dt until s!.sprite.frame == 10;
    {Sound::PingusSound::play_sound("plop")};

    // 3. particles, terrain, explosion sprite
    await game.dt until s!.sprite.frame == 13;
    spawn PinguParticles(<...>) in particles;
    call Game_Remove({&bomber_radius}, <...>);
    do
      <...>
      spawn Sprite(<...>);      // explosion

      // 4. tick: hides the explosion sprite
      await game.dt;
    end
    await FOREVER;
  end

  // 5. kills the pingu
  escape DEAD;
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ The \emph{Bomber action} sequence.
\label{lst.bomber}
}
\end{figure*}

The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius.%
\footnote{Bomber action animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#2} }
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber}) with actions associated to specific frames as
follows%
\footnote{State machine animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#3} }:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

Figure~\ref{lst.bomber} compares the implementations in C++ and \CEU.

In C++, the class \code{Bomber} defines the callbacks \code{draw} and
\code{update} to manage the state machine described above:
%
The class first defines one state variable for each action to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) updates the pingu animation and
movement on every frame regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
switches to \emph{state-2} (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 25--32 and
43--44).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 35--37).
%
Note that a single numeric state variable suffices to track the states, but the
original authors probably chose to encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., \emph{methods + mutable state}).
%
Like the C++ implementation for detecting double clicks, note also that the
state machine is encoded across 3 different methods, each intermixing code with
unrelated functionality.

The equivalent code for the \emph{Bomber action} in \CEU doesn't require state
variables and reflects the sequential state machine implicitly, using
\code{await} statements in direct style to separate the actions.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine~\cite{sync_async.cooperative}: a subroutine that retains runtime
state, such as local variables and the program counter, across reactions to
events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other abstractions
and execute in separate through the \code{spawn} primitive (ln. 4--5).
The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every frame.
%
The code tracks the animation instance (ln. 7--27), performing the last bomber
action on termination (ln. 30).
As soon as the animation starts, the code performs the first action (ln. 8).
The intermediate actions are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all actions happen in a contiguous
chunk of code (ln. 5--30) which handles no extra functionality.

@@

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{story-anim}
%\caption{The \emph{Story screen}.
%\label{fig.story}
%}
%\end{figure}

@@

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens%
\footnote{Story screen animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#4} }.
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.
%
Figure~\ref{lst.story} compares the implementations in C++ and \CEU.

@@

\textbf{Case Study: \emph{Bomber action} \code{draw} and \code{update} dispatching}

TODO

Figure~\ref{lst.hier} compares the implementations in C++ and \CEU.

@@

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ \emph{Bomber action} \code{draw} and \code{update} dispatching.
\label{lst.hier}
}
\end{figure*}

In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
animation frames (Figure~\ref{fig.bomber}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

@@

\textbf{Case Study: Managing the Pingus Lifecycle}

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{pingus_create_die-anim}
%\caption{Creation and death of pingus.
%\label{fig.pingus_create_die}
%}
%\end{figure}

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude%
\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
%
Figure~\ref{lst.pingus} compares the implementations in C++ and \CEU.

@@

code/await Pingu (<...>) do
    <...>
    loop do
        await game.dt;
        if Pingu_Is_Out_Of_Screen() then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Managing the pingus lifecycle.
\label{lst.pingus}
}
\end{figure*}

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen spawns a new \code{Pingu} on every invocation of
\code{Pingu\_Spawn}.
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

@@

%\begin{figure}[t]
%\centering
%\includegraphics[width=\columnwidth]{options-anim-opt}
%\caption{The \emph{Mouse Grab} configuration option.
%\label{fig.options}
%}
%\end{figure}

@@

\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ TODO.
\label{lst.grab}
}
\end{figure*}

The \emph{mouse grab option} restricts the mouse movement to the game window
boundaries%
\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
In addition, the \emph{Options menu} has a check box to toggle the
\emph{mouse grab option} with mouse clicks while still responding to
\emph{Ctrl-G} presses.
%
Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.
